---
title: "Microscopy Analysis of Cocoa Pollination and Gynoecium Fungal Infection"
subtitle: "Pollination Exclusion Experiment Across a Deforestation Gradient in Ghana"
author: "Nikita Maksimov"
date: "`r Sys.Date()`"
output:
html_document:
  theme: flatty
  toc: true
  toc_float: true
  number_sections: true
  df_print: paged
# bibliography: references.bib
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(readxl)

library(tidyr)
library(dplyr)
library(purrr)

library(ggplot2)

library(knitr)

# Statistical testing
library(lme4)

# library(MASS) # for negative binomial / it has select() function which typically conflict with select from dplyr

# conflicts()
# get("select")
# detach("package:MASS", unload = TRUE)
```

## Data import and cleaning

```{r read_and_clean, echo=FALSE, warning=FALSE, message=FALSE}
# ---- Read sample metadata dataframe (all samples including metabarcoding controls) ----
# Reusable for microbiome and microscopy analyses
cocoa_sample_metadata <- read_xlsx(
        here("data", "sample_metadata_with_microscopy.xlsx"),
        sheet = "metadata",
        col_types = c("text","text","text","text","text",
                      "text", "date", "date", "date", "text",
                      "text", "numeric", "date", "date", "text",
                      "text", "numeric", "date", "logical", "text",
                      "numeric", "logical", "logical", "numeric", "logical",
                      "numeric", "logical", "numeric", "logical", "logical",
                      "numeric")
)  |>
        # Remove qiime2:types row (QIIME2 metadata requirement)
        slice(-1) |>
        # Remove hyphen in sample-id (QIIME2 metadata requirement)
        rename(sample_id = `sample-id`) 

# nrow(cocoa_sample_metadata)
# 336

# ---- Define microscopy annotation table for biological flower samples (microscopy variables as columns) ----
microscopy_annotation_df <- cocoa_sample_metadata |>
        dplyr::select(
                sample_id,
                sample_type,
                management_type,
                farm_id,
                tree_id,
                collection_batch,
                is_specimen,
                dye,
                dye_concentration,
                specimen_qc_pass,
                is_pollination_dic,
                pi_dic,
                is_pollination_clsm,
                pi_clsm,
                is_germination,
                callose_plug_index,
                is_hyphae_present,
                is_spore_present,
                fungal_colonization_score) |>
        filter(sample_type %in% c("bagged_flower", "unbagged_flower")) |> 
        # For semantic clarity since metabarcoding controls are removed
        rename(flower_treatment = sample_type) |> 
        # Factorize categorical variable
        mutate(
                across(
                        c(flower_treatment,
                          management_type,
                          farm_id,
                          tree_id,
                          collection_batch,
                          dye),
                        as.factor
                ),
                # Infection score is an ordered factor
                fungal_colonization_score = factor(
                        fungal_colonization_score,
                        levels = 0:5,
                        ordered = TRUE
                )
        )

# nrow(microscopy_annotation_df)
# 294

# ---- Define analysis-ready microscopy dataset (QC-passed specimens only) ----
microscopy_df <- microscopy_annotation_df |> 
        filter(specimen_qc_pass == TRUE)

# nrow(microscopy_df)
# 196
```

## Data structure overview

The `microscopy_annotation_df` contains all flower samples observations of cocoa flower stigmas in a pollination exclusion experiment. Each row corresponds to a single flower used for DNA extraction for the metabarcoding experiment. Some rows are populated with microscopy data. The `microscopy_df` is a subset including only specimens that passed microscopy quality control and were successfully phenotyped. The analysis focuses on pollination phenotypes and fungal infection traits of the stigmas and includes following variables:

-   **`sample_id`** - character; unique sample identifier matching QIIME2 naming (pattern `[cp][a-z]{2}[1-7][1-3]`).
-   **`flower_treatment`** - factor with 2 levels: `"bagged_flower"` and `"unbagged_flower"`. Indicates whether a flower was accessible to insect visitors capable of pollination. The first character of the `sample_id` (c or p) encodes the flower treatment, corresponding to bagged and unbagged flowers, respectively.
-   **`management_type`** - factor with 4 levels: `"agroforest"`, `"full_sun"`, `"inside_forest"`, `"near_forest"`. Describes the land-use context of the sampling site.
-   **`farm_id`** - factor with 7 levels; unique farm identifier (patter [a-z]{2}).
-   **`tree_id`** - factor with 49 levels; unique tree identifier (pattern [a-z]{2}[1-7].
-   **`collection_batch`** - date; field collection batch corresponding to sampling date.
-   **`is_specimen`** logical; whether a microscopy specimen was successfully prepared. All prepared specimens were analysed under dark field differential interference contrast (DIC) stereo microscope.
-   **`specimen_qc_pass`** - logical; indicates whether the specimen passed the quality control based on DIC microscopy and was included in further analysis. Specimens were included if all five stigma lobes were present on the slide. Mechanical fragmentation during squashing was allowed, provided that the entire receptive surface was retained. All specimens passing this quality control were subsequently imaged by confocal laser scanning microscopy (CLSM).
-   **`dye`** - factor with 3 levels; most samples were stained with decolourised aniline blue (`"dAB"`) dye with affinity to callose (pollen tube cell wall polysacharide) and lower affinity to chitin (major component of fungal cell walls). Ruthenium red (`"RR"`) was used in few trial experiments to visualize pollen grains, but have been considered inefficient. Several `"no_stain"` specimens were prepared as negative controls. Pollen performance (germination) and fungal infection were quantified only on dAB stained specimens.
-   **`dye_concentration`** - numeric; 0.1% dAB was the major most efficient staining method. Lower concentration (0.01%) was used in few trial experiments.
-   **`is_pollination_dic`** - logical; pollen presence based on DIC microscopy. At least one pollen grain was sufficient to classify a specimen as pollinated.
-   **`pi_dic`** - numeric; pollination intensity measured by DIC microscopy, expressed as the number of pollen grains per stigma. Values are NA when no pollen was detected or no specimen available.
-   **`is_pollination_clsm`** - logical; presence of pollen grains on the stigma assessed by confocal laser scanning microscopy (CLSM).
-   **`pi_clsm`** - numeric; pollination intensity measured by CLSM, expressed as the number of pollen grains per stigma. Pollen grains were counted manualy on z-stack images in Zen Lite. Values are NA when pollen was not detected or specimen did not pass the quality control.
-   **`is_germination`** - logical; indicates the presence of pollen tubes in stigma/style by fluorscence from callose plugs, assessed by CLSM imaging. In few cases pollen grains were absent on the stigma surface but strong callose plug fluorscense was detected indicating washing out of pollen grains during specimen preparation.
-   **`callose_plug_index`** - numeric. Quantitative index describing the abundance of callose plugs in region of interest in the focal plane with maximal number of callose plugs. It was obtained using custom Python scikit-image script for Gaussian of  Laplacian blob detection algorithm on tile CLSM images. It was assessed only in dAB stained specimens. Few specimens with visible pollen tubes in style canal had no pollen grains on the stigma surface (pollen grains might be washed out during DNA extraction or during specimen preparation).
-   **`is_hyphae_present`**  logical; indicates the presence of fungal hyphae on the stigma surface, assessed by CLSM only for dAB stained specimens.
-   **`is_spore_present`** - logical. indicates the presence of fungal spores on the stigma surface. This variable is sparse and used for
descriptive purposes only.
-   **`fungal_colonization_score`** - ordered factor. Rank-based score describing how many stigma lobes have fungal hyphae, ranging from no colonization (0) to strong colonization (5).

```{r data_structure_overview, echo=FALSE}
# ---- Filtered microscopy data structure overview ----
str(microscopy_annotation_df)

# ---- List of levels for all factor variables ----
microscopy_annotation_df |>
        dplyr::select(where(is.factor)) |>
        map(levels)
```

## Data inspection

```{r inspect_data, echo=FALSE, results="hide", message=FALSE, warning=FALSE}
# ---- All specimens were assigned status based on DIC microscopy ----
identical(
        microscopy_annotation_df |> filter(is_specimen == TRUE),
        microscopy_annotation_df |> filter(!is.na(specimen_qc_pass))
)

# ---- All successful specimens were assessed for pollination ----
# by DIC microscopy:
identical(
        microscopy_annotation_df |> filter(specimen_qc_pass == TRUE),
        microscopy_annotation_df |> filter(!is.na(is_pollination_dic))
)

# by CLSM:
identical(
        microscopy_annotation_df |> filter(specimen_qc_pass == TRUE),
        microscopy_annotation_df |> filter(!is.na(is_pollination_clsm))
)

# ---- Pollination intensity is defined only for pollinated specimens (DIC) ----
all(
        microscopy_annotation_df |>
                filter(is_pollination_dic == FALSE) |>
                pull(pi_dic) |>
                is.na()
)

# ---- Pollination intensity is defined only for pollinated specimens (CLSM) ----
all(
        microscopy_annotation_df |>
                filter(is_pollination_clsm == FALSE) |>
                pull(pi_clsm) |>
                is.na()
)

# ---- Presence of pollination intensity implies pollination (CLSM) ----
all(
        microscopy_annotation_df |>
                filter(!is.na(pi_clsm)) |>
                pull(is_pollination_clsm)
)

# ---- Germination can only be assessed in pollinated specimens (CLSM) ----
all(
        microscopy_annotation_df |>
                filter(is_germination == TRUE) |>
                pull(is_pollination_clsm)
)

# ---- Fungal hyphae are assessed only in dAB-stained specimens ----
all(
        microscopy_annotation_df |>
                filter(is_hyphae_present == TRUE) |>
                pull(dye) == "dAB"
)


# ---- Not all QC-passed specimens were assessed for fungal structure presence ----
# Specimens with missing fungal phenotypes correspond exclusively to unstained microscopy controls and specimens stained with Ruthenium Red (RR)
all(
        microscopy_annotation_df |>
                filter(specimen_qc_pass == TRUE, is.na(is_hyphae_present)) |>
                pull(dye) %in% c("RR", "no_stain")
)

```


```{r set_up_global_okabe_ito_color_pallete, results='hide', echo=FALSE}
# ---- Global color palettes ----
okabe_ito <- c(
  oi_orange  = "#E69F00",
  oi_skyblue = "#56B4E9",
  oi_bluish  = "#0072B2",
  oi_green   = "#009E73",
  oi_vermilion = "#D55E00",
  oi_purple  = "#CC79A7",
  oi_grey    = "#999999",
  oi_light_grey = "#BDBDBD"
)

pal_microscopy_status <- c(
  "Microscopy data available" = unname(okabe_ito["oi_bluish"]),
  "Incomplete specimen"       = unname(okabe_ito["oi_orange"]),
  "No specimen"               = unname(okabe_ito["oi_light_grey"])
)

pal_pollination <- c(
  "Pollinated"     = unname(okabe_ito["oi_green"]),
  "Not pollinated" = unname(okabe_ito["oi_skyblue"])
)

pal_germination <- c(
  "Flowers with germinated pollen"     = unname(okabe_ito["oi_vermilion"]),
  "Flowers without germinated pollen" = unname(okabe_ito["oi_grey"])
)

pal_flower_treatment <- c(
  "Bagged flower"            = unname(okabe_ito["oi_bluish"]),
  "Unbagged flower" = unname(okabe_ito["oi_vermilion"])
)

pal_hyphae <- c(
  "Hyphae present" = unname(okabe_ito["oi_purple"]),
  "No hyphae"      = unname(okabe_ito["oi_grey"])
)

microscopy_status_theme <- theme_minimal(base_size = 12) + 
        theme(plot.title = element_text(size = 14, hjust = 0.5),
              plot.subtitle = element_text(size = 12, hjust = 0.5),
              legend.position = "bottom" )

```


## Exploring microscopy data availability by sampling groups within the metabarcoding dataset

Microscopy data availability was visualized across key experimental grouping variables (`flower_treatment`, `management_type` and `farm_id`) to evaluate class imbalance in the microscopy subset.
In total, microscopy data were available for 196 flowers out of 294 flowers included in the metabarcoding experiment. Of these, 46% (90 samples) correspond to the `bagged_flower` treatment, whereas 54% (106 samples) belong to the `unbagged_flower` treatment.

```{r microscopy_data_availability_by_flower_treatment, echo=FALSE}
# ---- Microscopy data availability by flower treatment ----
microscopy_annotation_df |>
        mutate(
                flower_treatment = factor(
                        flower_treatment,
                        levels = c("bagged_flower", "unbagged_flower")
                ),
                microscopy_status = case_when(
                        specimen_qc_pass == TRUE                         ~ "Microscopy data available",
                        is_specimen == TRUE & specimen_qc_pass == FALSE  ~ "Incomplete specimen",
                        is_specimen == FALSE                             ~ "No specimen"
                )
        ) |>
        count(flower_treatment, microscopy_status) |>
        group_by(flower_treatment) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        mutate(
                microscopy_status = factor(
                        microscopy_status,
                        levels = c("Microscopy data available", "Incomplete specimen", "No specimen")
                )
        ) |>
        ggplot(aes(x = flower_treatment, y = percent, fill = microscopy_status)) +
        geom_col(
                width = 0.6, 
                color = "white", 
                position = position_stack(reverse = TRUE)
                 ) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
        ) +
        scale_y_continuous(
                labels = scales::percent_format(accuracy = 1),
                expand = c(0, 0)
        ) +
        scale_fill_manual(
                name = "",
                values = pal_microscopy_status
        ) +
        labs(
                title = "Microscopy data availability by flower treatment",
                x = "Flower treatment",
                y = "Percentage of flowers"
        ) +
        microscopy_status_theme

ggsave(
  filename = "results/Нmicroscopy_data_avail_by_flower_treatment.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)
```
Across all management types, more than half of the flower samples were associated with successfully acquired microscopy data.
```{r microscopy_data_availability_by_management_type, echo=TRUE}
# ---- Microscopy data availability by management type ----
microscopy_annotation_df |>
        mutate(
                management_type = factor(
                        management_type,
                        levels = c("inside_forest", "near_forest", "agroforest", "full_sun")
                ),
                microscopy_status = case_when(
                        specimen_qc_pass == TRUE                         ~ "Microscopy data available",
                        is_specimen == TRUE & specimen_qc_pass == FALSE  ~ "Incomplete specimen",
                        is_specimen == FALSE                             ~ "No specimen"
                )
        ) |>
        count(management_type, microscopy_status) |>
        group_by(management_type) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        mutate(
                microscopy_status = factor(
                        microscopy_status,
                        levels = c("Microscopy data available", "Incomplete specimen", "No specimen")
                )
        ) |>
        ggplot(aes(x = management_type, y = percent, fill = microscopy_status)) +
        geom_col(
                width = 0.7, 
                color = "white",
                position = position_stack(reverse = TRUE)
                ) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
        ) +
        scale_y_continuous(
                labels = scales::percent_format(accuracy = 1),
                expand = c(0, 0)
        ) +
        scale_fill_manual(
                name = "",
                values = pal_microscopy_status
        ) +
        labs(
                title = "Microscopy data availability by management type",
                x = "Management type",
                y = "Percentage of flowers"
        ) +
        microscopy_status_theme

ggsave(
  filename = "results/microscopy_data_avail_by_management.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)

```
Similarly, for each individual farm, microscopy data were available for more than 50% of the collected flower samples, although the proportion of incomplete or missing specimens varied among farms.

```{r microscopy_data_availability_by_farm_id, echo=TRUE}
# ---- Microscopy data availability by farms ----
microscopy_annotation_df |>
        mutate(
                management_type = factor(
                        management_type,
                        levels = c("inside_forest", "near_forest", "agroforest", "full_sun")
                ),
                farm_id = factor(
                        farm_id,
                        levels = microscopy_annotation_df |>
                                distinct(farm_id, management_type) |>
                                arrange(management_type, farm_id) |>
                                pull(farm_id)
                ),
                microscopy_status = case_when(
                        specimen_qc_pass == TRUE                         ~ "Microscopy data available",
                        is_specimen == TRUE & specimen_qc_pass == FALSE  ~ "Incomplete specimen",
                        is_specimen == FALSE                             ~ "No specimen"
                )
        ) |>
        count(farm_id, microscopy_status) |>
        group_by(farm_id) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        mutate(
                microscopy_status = factor(
                        microscopy_status,
                        levels = c("Microscopy data available", "Incomplete specimen", "No specimen")),
                # Explicit levels to order by management type
                # inside_forest (ib) => near_forest (vr, sa) => agroforest (kk, mt) => full_sun (vi, yb)
                farm_id = factor(
                        farm_id,
                        levels = c("ib", "vr", "sa", "kk", "mt", "vi", "yb"))
        ) |>
        ggplot(aes(x = farm_id, y = percent, fill = microscopy_status)) +
        geom_col(
                width = 0.7,
                color = "white",
                position = position_stack(reverse = TRUE)
                ) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
        ) +
        scale_y_continuous(
                labels = scales::percent_format(accuracy = 1),
                expand = c(0, 0)
        ) +
        scale_fill_manual(
                name = "",
                values = pal_microscopy_status
        ) +
        labs(
                title = "Microscopy data availability by farm",
                x = "Farm ID",
                y = "Percentage of flowers"
        ) +
        microscopy_status_theme

ggsave(
  filename = "results/microscopy_data_avail_by_farm.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)

```

Microscopy data were obtained for at least one flower from all 49 trees sampled for the metabarcoding experiment. To assess whether comparisons between flower treatments are balanced at the tree level, treatment representation was evaluated for each tree in the microscopy subset. Overall, 80% of trees (39 out of 49) were represented by at least one `bagged_flower` and one `unbagged_flower`, whereas 20% of trees (10 out of 49) were represented by only a single `flower_treatment`.

```{r microscopy_pairing_summary, echo=FALSE}
microscopy_df |>
        distinct(tree_id, flower_treatment) |>
        count(tree_id, name = "n_treatments") |>
        count(n_treatments, name = "n_trees") |>
        mutate(
                flower_treatment_representation = if_else(
                        n_treatments == 2,
                        "Both flower treatments",
                        "Single flower treatment"
                        )
                ) |>
        dplyr::select(flower_treatment_representation, n_trees)

```

This 10 trees contributed to 22 flower from the same `flower_treatment` group. This imbalance was not randomly distributed among `management_type` and `farm_id`. Unpaired trees were concentrated within a single `inside_forest` farm (`ib`) and, to a lesser extent, within one `near_forest` farm (`vr`), reflecting differential loss of microscopy specimens during preparation and screening.

```{r unpaired_tree_structure, echo=FALSE}
# ---- Structure of trees with unpaired flower_treatment and corresponding microscopy samples ----
microscopy_df |>
        distinct(tree_id, flower_treatment) |>
        count(tree_id, name = "n_treatments") |>
        filter(n_treatments == 1) |>
        dplyr::select(tree_id) |>
        left_join(
                microscopy_df |>
                        dplyr::select(
                                tree_id,
                                management_type,
                                farm_id,
                                flower_treatment,
                                sample_id
                                ),
                by = "tree_id"
                ) |>
        group_by(tree_id, management_type, farm_id, flower_treatment) |>
        summarise(
                n_microscopy_samples = n(),
                microscopy_samples = paste(sample_id, collapse = ", "),
                .groups = "drop"
                ) |>
        arrange(management_type, farm_id, tree_id)

```

These patterns indicate that variation in microscopy data availability is partly structured at the farm and tree levels. To account for this structure, downstream analyses consider both the full microscopy dataset and a more stringent subset restricted to trees represented by both flower treatments (at least one flower per each `flower_treatment`).

```{r microscopy_paired_subset, echo=FALSE}
microscopy_paired_df <- microscopy_df |>
        group_by(tree_id) |>
        filter(n_distinct(flower_treatment) == 2) |>
        ungroup()
```

#### Pollination outcome by microscopy method in whole metabarcoding dataset

```{r pollination_outcome_by_microscopy_method, echo=FALSE}
# ---- Horizontal stacked barplot: pollination outcome by microscopy method ----
bind_rows(
  tibble(
    Method   = "DIC",
    Category = c("No specimen", "Incomplete specimen", "Not pollinated", "Pollinated"),
    Count    = c(
      sum(microscopy_annotation_df$is_specimen != TRUE, na.rm = TRUE),
      sum(microscopy_annotation_df$specimen_qc_pass == FALSE, na.rm = TRUE),
      sum(microscopy_annotation_df$is_pollination_dic == FALSE, na.rm = TRUE),
      sum(microscopy_annotation_df$is_pollination_dic == TRUE,  na.rm = TRUE)
    )
  ),
  tibble(
    Method   = "CLSM",
    Category = c("No specimen", "Incomplete specimen", "Not pollinated", "Pollinated"),
    Count    = c(
      sum(microscopy_annotation_df$is_specimen == FALSE, na.rm = TRUE),
      sum(microscopy_annotation_df$specimen_qc_pass == FALSE, na.rm = TRUE),
      sum(microscopy_annotation_df$is_pollination_clsm == FALSE, na.rm = TRUE),
      sum(microscopy_annotation_df$is_pollination_clsm == TRUE,  na.rm = TRUE)
    )
  )
) |>
  group_by(Method) |>
  mutate(
    Percent = round(100 * Count / sum(Count), digits = 1),
    Category = factor(
      Category,
      levels = c(
        "Pollinated",
        "Not pollinated",
        "Incomplete specimen",
        "No specimen"
      )
    )
  ) |>
  ungroup() |>
  ggplot(aes(x = Method, y = Percent, fill = Category)) +
  geom_col(width = 0.6, color = "white", position = position_stack(reverse = TRUE)) +
  geom_text(
    aes(label = Count),
    position = position_stack(vjust = 0.5, reverse = TRUE),
    size = 3,
  ) +
  scale_fill_manual(
    values = c(
      "No specimen"           = "#CCCCCC",
      "Incomplete specimen"   = "#999999",
      "Not pollinated"        = "#56B4E9",
      "Pollinated"            = "#009E73"
    ),
    guide = guide_legend(),
    name = NULL
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent_format(scale = 1),
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    title = "Pollination outcome by microscopy method",
    x = NULL,
    y = "Percentage of flowers included in metabarcoding"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.y = element_text(face = "bold"),
    legend.position = "bottom"
  )

ggsave(
  filename = "results/stacked_barplot_DICvsCLSM_pollination.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)

```

## Comparing CLSM and dark field DIC microscopy for pollination analysis in cocoa

DIC microscopy underestimates pollination compared to CLSM on the same specimen set.

```{r clsm_vs_dic_for_pollination_detection, echo=FALSE}
# ---- Pollination detection by microscopy method (stacked barplots) ----
bind_rows(
        microscopy_df |>
                filter(!is.na(is_pollination_dic)) |>
                transmute(
                        Method = "DIC",
                        Pollination_status = if_else(
                                is_pollination_dic,
                                "Pollinated",
                                "Not pollinated"
                                )
                        ),
        microscopy_df |>
                filter(!is.na(is_pollination_clsm)) |>
                transmute(
                        Method = "CLSM",
                        Pollination_status = if_else(
                                is_pollination_clsm,
                                "Pollinated",
                                "Not pollinated"
                                )
                        )
        ) |>
        count(Method, Pollination_status) |>
        group_by(Method) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        mutate(
                Pollination_status = factor(
                        Pollination_status,
                        levels = c("Pollinated", "Not pollinated")
                        ),
                Method = factor(Method, levels = c("DIC", "CLSM"))
                ) |>
        ggplot(aes(x = Method, y = percent, fill = Pollination_status)) +
        geom_col(width = 0.6, color = "white", position = position_stack(reverse = TRUE)) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(
                labels = scales::percent_format(accuracy = 1),
                expand = c(0, 0)
                ) +
        scale_fill_manual(
                name = "",
                values = pal_pollination
                ) +
        labs(
                title = "Pollination detection by microscopy method",
                x = "Microscopy method",
                y = "Percentage of flowers"
                ) +
        theme_minimal(base_size = 12) +
        theme(
                plot.title = element_text(hjust = 0.5),
                legend.position = "bottom"
                )
ggsave(
  filename = "results/pollination_by_microscopy_method.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)
```

Dark field DIC microscopy systematically underestimates pollination intensity, with increasing deviation at higher pollen loads. DIC microscopy could be used as more available and cheap alternative for evaluation of pollination and estimation of pollination intensity. For downstream analysis only CLSM data was used.

```{r dic_vs_clsm_for_pollination_intensity, echo=FALSE}
# ---- Relationship between DIC and CLSM pollination intensity ----
microscopy_df |>
        filter(
                !is.na(pi_dic),
                !is.na(pi_clsm)
                ) |>
        ggplot(aes(x = pi_dic, y = pi_clsm)) +
        geom_point(size = 2, alpha = 0.6) +
        geom_smooth(
                method = "lm",
                se = FALSE,
                color = "black",
                linewidth = 0.8
                ) +
        geom_abline(
                intercept = 0,
                slope = 1,
                color = "grey60"
                ) +
        annotate(
                "text",
                x = 170, y = 300,
                label = "Linear regression line",
                hjust = 0,
                size = 3
                ) +
        annotate(
                "text",
                x = 220, y = 200,
                label = "Perfect agreement line",
                hjust = 0,
                size = 3
                ) +
        labs(title = "Comparison of dark field differential interference contrast (DIC)\nand confocal laser scanning microscopy (CLSM) methods \nfor pollination intensity analysis",
             x = "Pollination intensity (DIC microscopy), [pollen grains/stigma]",
             y = "Pollination intensity (CLSM), [pollen grains/stigma]"
             ) + 
        coord_cartesian(xlim = c(0, 300), ylim = c(0, 400)) +
        theme_minimal(base_size = 12) +
        theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
              panel.grid.major = element_line(color = "grey85", linewidth = 0.4), 
              panel.grid.minor = element_blank(),
              plot.title = element_text(hjust = 0.5),
              legend.position = "bottom"
              )
```

## Pollination phenotyping
### Proportion of pollinated flowers across farms

```{r pollination_by_farm_unbagged_flowers, echo=FALSE}
# ---- Pollination outcome by farm and management type (only unbagged flowers) ----
microscopy_df |>
        filter(
                flower_treatment == "unbagged_flower",
                !is.na(is_pollination_clsm)
                ) |>
        mutate(
                farm_id = factor(
                        farm_id,
                        levels = c("ib", "vr", "sa", "kk", "mt", "vi", "yb")
                        ),
                pollination_status = if_else(
                        is_pollination_clsm,
                        "Pollinated",
                        "Not pollinated"
                        )
                ) |>
        mutate(
                pollination_status = factor(
                        pollination_status,
                        levels = c("Pollinated", "Not pollinated")
                        )
                ) |> 
        mutate(
                management_type = factor(
                        management_type,
                        levels = c(
                                "inside_forest",
                                "near_forest",
                                "agroforest",
                                "full_sun"
                                )
                        )
                ) |> 
        count(management_type, farm_id, pollination_status) |>
        group_by(farm_id) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = farm_id, y = percent, fill = pollination_status)) +
        geom_col(width = 0.7, color = "white", position = position_stack(reverse = TRUE)) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
        scale_fill_manual(
                name = "",
                values = pal_pollination
                ) +
        facet_grid(
                . ~ management_type,
                scales = "free_x",
                space  = "free_x",
                switch = "x"
                ) +
        labs(
                title = "Proportion of pollinated flowers across farms",
                subtitle = "(unbagged flowers only)",
                x = "Farm ID and management type",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme +
        theme(
                strip.placement = "outside",
                strip.background = element_blank(),
                axis.text.x = element_text(margin = margin(t = 5))
                )
ggsave(
  filename = "results/proportion_of_pollinated_flowers_across_farms_unbagged_flowers_only.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)

```


```{r pollination_by_farm_both_flower_treatments, echo=FALSE}
# ---- Pollination outcome by farm and management type (both bagged and unbagged flowers) ----
microscopy_df |>
        filter(
                !is.na(is_pollination_clsm)
                ) |>
        mutate(
                farm_id = factor(
                        farm_id,
                        levels = c("ib", "vr", "sa", "kk", "mt", "vi", "yb")
                        ),
                pollination_status = if_else(
                        is_pollination_clsm,
                        "Pollinated",
                        "Not pollinated"
                        )
                ) |>
        mutate(
                pollination_status = factor(
                        pollination_status,
                        levels = c("Pollinated", "Not pollinated")
                        )
                ) |> 
        mutate(
                management_type = factor(
                        management_type,
                        levels = c(
                                "inside_forest",
                                "near_forest",
                                "agroforest",
                                "full_sun"
                                )
                        )
                ) |> 
        count(management_type, farm_id, pollination_status) |>
        group_by(farm_id) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = farm_id, y = percent, fill = pollination_status)) +
        geom_col(width = 0.7, color = "white", position = position_stack(reverse = TRUE)) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
        scale_fill_manual(
                name = "",
                values = pal_pollination
                ) +
        facet_grid(
                . ~ management_type,
                scales = "free_x",
                space  = "free_x",
                switch = "x"
                ) +
        labs(
                title = "Proportion of pollinated flowers across farms",
                subtitle = "(both bagged and unbagged flowers)",
                x = "Farm ID and management type",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme +
        theme(
                strip.placement = "outside",
                strip.background = element_blank(),
                axis.text.x = element_text(margin = margin(t = 5))
                )

ggsave(
  filename = "results/proportion_of_pollinated_flowers_across_farms_bagged_and_unbagged_flowers.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)

```

### Effectiveness of bagging as pollinator exclusion method
Pollen detection is consistently lower in `bagged_flower`s compared to `unbagged_flower`s across farms and management types. While this pattern suggests an effect of pollinator exclusion, formal inference requires accounting for the non-independence of flowers sampled from the same tree.
```{r proportion_of_pollinated_flowers_by_flower_treatment_clsm, echo=FALSE}
# ---- Proportion of pollinated flowers by treatment (all flowers with available microscopy data) ----
microscopy_df |>
        filter(!is.na(is_pollination_clsm)) |>
        mutate(
                flower_treatment = factor(
                        flower_treatment,
                        levels = c("bagged_flower", "unbagged_flower")
                        ),
                pollination_status = if_else(is_pollination_clsm, "Pollinated", "Not pollinated"),
                pollination_status = factor(pollination_status, levels = c("Pollinated", "Not pollinated"))
                ) |>
        count(flower_treatment, pollination_status, name = "n") |>
        group_by(flower_treatment) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = flower_treatment, y = percent, fill = pollination_status)) +
        geom_col(width = 0.7, color = "white", position = position_stack(reverse = TRUE)) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(
                labels = scales::percent_format(accuracy = 1),
                expand = c(0, 0)
                ) +
        scale_fill_manual(
                name = "",
                values = pal_pollination,
                breaks = c("Pollinated", "Not pollinated")
                ) +
        labs(
                title = "Proportion of pollinated flowers by treatment",
                subtitle = "(all flowers with available microscopy data)",
                x = "Flower treatment",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme

ggsave(
  filename = "results/proportion_of_flowers_by_flower_treatment_all_flowers.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)
```

These stack barplot includes data only from trees with at least one bagged and unbagged flower with available microscopy data.

```{r proportion_of_pollinated_flowers_by_flower_treatment_clsm_strict, eval=FALSE}
# ---- Proportion of pollinated flowers (CLSM) by treatment (only trees with at least one bagged and unbagged flower) ----
microscopy_paired_df |>
        filter(!is.na(is_pollination_clsm)) |>
        mutate(
                flower_treatment = factor(
                        flower_treatment,
                        levels = c("bagged_flower", "unbagged_flower")
                        ),
                pollination_status = if_else(is_pollination_clsm, "Pollinated", "Not pollinated"),
                pollination_status = factor(pollination_status, levels = c("Pollinated", "Not pollinated"))
                ) |>
        count(flower_treatment, pollination_status, name = "n") |>
        group_by(flower_treatment) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = flower_treatment, y = percent, fill = pollination_status)) +
        geom_col(width = 0.7, color = "white", position = position_stack(reverse = TRUE)) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(
                labels = scales::percent_format(accuracy = 1),
                expand = c(0, 0)
                ) +
        scale_fill_manual(
                name = "",
                values = pal_pollination,
                breaks = c("Pollinated", "Not pollinated")
                ) +
        labs(
                title = "Proportion of pollinated flowers by treatment",
                subtitle = "(only trees with at least one bagged and unbagged flower)",
                x = "Flower treatment",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme

```

Pollination frequency differed significantly between flower treatments. After accounting for non-independence of flowers sampled from the same tree using a binomial generalized linear mixed model, unbagged flowers showed a substantially higher probability of pollen detection compared to bagged flowers (odds ratio в‰€ 4.1, p < 0.001). Inclusion of farm identity as an additional random effect did not improve model fit, as between-farm variance was negligible after accounting for tree-level variation. Therefore, subsequent analyses retained tree identity as the sole random effect.

```{r binomial_glmm_pollination_flower_treatment, echo=FALSE}

# H0: The probability of detecting pollen is independent of whether a flower was bagged or unbagged.
# H1: The probability of detecting pollen is association with flower treatment. 

glmm_pollination_flower_treatment <- glmer(
  is_pollination_clsm ~ flower_treatment + (1 | tree_id),
  data = microscopy_df,
  family = binomial
)

#print(glmm_pollination_flower_treatment)

summary(glmm_pollination_flower_treatment)
```


```{r}
library(performance)

performance::r2_nakagawa(glmm_pollination_flower_treatment)
```


### Pollination intensity

Residual pollination in control group (**`bagged_flower`**) might be caused by insects able to get through the bag or insects preexisting in flower buds prior to bagging. Bagged flowers show predominantly low pollination intensity with rare high-intensity events. Only one sample in control group had seed saturating pollination intensity (\> 110 pollen grains/stigma) (M. Falque et al., 1995). To facilitate downstream microbiome analyses, pollination intensity was treated as an ordinal proxy for pollinator visitation, distinguishing marginal pollen presence from moderate and saturating pollen loads


Pollination intensity distribution is zero-inflated (many non-pollinated flowers) and highly skewed (long tail with rare successful pollination events).
```{r distribution_of_pi_clsm_by_flower_treatment, echo=FALSE}
microscopy_df |>
        filter(!is.na(pi_clsm)) |>
        filter(pi_clsm > 0) |> # This is zero truncated data NB!
        ggplot(aes(x = pi_dic)) +
        geom_histogram(binwidth = 20, color = "black", alpha = 0.7) +
        geom_rug(sides = "bl") +
        geom_vline(xintercept = 238, linetype = "dashed", linewidth = 0.6) +
        geom_vline(xintercept = 43,  linetype = "dotted", linewidth = 0.6) +
        facet_wrap(~ flower_treatment, ncol = 1) +
        labs(
                title = "Distributions of pollination intensity by flower treatment",
                subtitle = paste(
                        "Reference thresholds from Falque et al. (1995):\n",
                        "dotted = 50% of maximum fruit set (~43 pollen grains)\n",
                        "dashed = seed-saturating pollination (~238 pollen grains)"),
                x = "Pollination intensity, [pollen grains/stigma]",
                y = "Number of pollinated flowers"
        ) +
        microscopy_status_theme

ggsave(
  filename = "results/distribution_of_pi_clsm_by_flower_treatment.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)
```

### Pollen performance (pollen germination and callose plug index)

###

```{r proportion_of_flowers_with_germinated_pollen_by_flower_treatment, echo=FALSE}
# ---- Proportion of flowers with germinated pollen by flower treatment (all flowers with available microscopy data) ----
microscopy_df |>
        filter(!is.na(is_germination)) |>
        mutate(
                flower_treatment = factor(
                        flower_treatment,
                        levels = c("bagged_flower", "unbagged_flower")
                        ),
                germination_status = if_else(is_germination, "Flowers with germinated pollen", "Flowers without germinated pollen"),
                germination_status = factor(germination_status, levels = c("Flowers with germinated pollen", "Flowers without germinated pollen"))
                ) |>
        count(flower_treatment, germination_status, name = "n") |>
        group_by(flower_treatment) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = flower_treatment, y = percent, fill = germination_status)) +
        geom_col(width = 0.7, color = "white", position = position_stack(reverse = TRUE)) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(
                labels = scales::percent_format(accuracy = 1),
                expand = c(0, 0)
                ) +
        scale_fill_manual(
                name = "",
                values = pal_germination,
                breaks = c("Flowers with germinated pollen", "Flowers without germinated pollen")
                ) +
        labs(
                title = "Proportion of flowers with germinated pollen by flower treatment",
                subtitle = "(all flowers with available microscopy data)",
                x = "Flower treatment",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme

ggsave(
  filename = "results/proportion_of_flowers_with_germinated_pollen_by_flower_treatment.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)
```

### Proportion of flowers with germinated pollen across farms

Flowers with callose plug signal were treated as having germinated pollen grains. The following visualization represents percentage of such flowers across farms ordered by management type.  

```{r pollen_germination_by_farm_unbagged_flowers, echo=FALSE}
# ---- Flowers with germinated pollen by farm and management type (only unbagged flowers) ----
microscopy_df |>
        filter(
                flower_treatment == "unbagged_flower",
                !is.na(is_germination)
                ) |>
        mutate(
                farm_id = factor(
                        farm_id,
                        levels = c("ib", "vr", "sa", "kk", "mt", "vi", "yb")
                        ),
                germination_status = if_else(
                        is_germination,
                        "Flowers with germinated pollen",
                        "Flowers without germinated pollen"
                        )
                ) |>
        mutate(
                germination_status = factor(
                        germination_status,
                        levels = c("Flowers with germinated pollen", "Flowers without germinated pollen")
                        )
                ) |> 
        mutate(
                management_type = factor(
                        management_type,
                        levels = c(
                                "inside_forest",
                                "near_forest",
                                "agroforest",
                                "full_sun"
                                )
                        )
                ) |> 
        count(management_type, farm_id, germination_status) |>
        group_by(farm_id) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = farm_id, y = percent, fill = germination_status)) +
        geom_col(width = 0.7, color = "white", position = position_stack(reverse = TRUE)) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
        scale_fill_manual(
                name = "",
                values = pal_germination
                ) +
        facet_grid(
                . ~ management_type,
                scales = "free_x",
                space  = "free_x",
                switch = "x"
                ) +
        labs(
                title = "Proportion of flowers with germinated pollen across farms\n(unbagged flowers only)",
                x = "Farm ID and management type",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme +
        theme(
                strip.placement = "outside",
                strip.background = element_blank(),
                axis.text.x = element_text(margin = margin(t = 5))
                )
ggsave(
  filename = "results/proportion_of_flowers_with_germinated_pollen_across_farms_unbagged_only.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)

```

The same stacked barplot based on both treatment groups is distorted by the `bagged_flower`s with low pollination.

```{r germination_by_farm_both_flower_treatments, echo=FALSE}
# ---- Germination by farm and management type (both bagged and unbagged flowers) ----
microscopy_df |>
        filter(
                !is.na(is_germination)
                ) |>
        mutate(
                farm_id = factor(
                        farm_id,
                        levels = c("ib", "vr", "sa", "kk", "mt", "vi", "yb")
                        ),
                germination_status = if_else(
                        is_germination,
                        "Flowers with germinated pollen",
                        "Flowers without germinated pollen"
                        )
                ) |>
        mutate(
                germination_status = factor(
                        germination_status,
                        levels = c("Flowers with germinated pollen", "Flowers without germinated pollen")
                        )
                ) |> 
        mutate(
                management_type = factor(
                        management_type,
                        levels = c(
                                "inside_forest",
                                "near_forest",
                                "agroforest",
                                "full_sun"
                                )
                        )
                ) |> 
        count(management_type, farm_id, germination_status) |>
        group_by(farm_id) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = farm_id, y = percent, fill = germination_status)) +
        geom_col(width = 0.7, color = "white", position = position_stack(reverse = TRUE)) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
        scale_fill_manual(
                name = "",
                values = pal_germination
                ) +
        facet_grid(
                . ~ management_type,
                scales = "free_x",
                space  = "free_x",
                switch = "x"
                ) +
        labs(
                title = "Proportion of flowers with germinated pollen across farms\n(both bagged and unbagged flowers)",
                x = "Farm ID and management type",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme +
        theme(
                strip.placement = "outside",
                strip.background = element_blank(),
                axis.text.x = element_text(margin = margin(t = 5))
                )

ggsave(
  filename = "pollen_germination_proportion_across_farms_bagged_and_unbagged_flowers.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)

```

# 

Scatter plot between pollination intensity and callose plug index. Flowers with high callose plug index and low pollination intensity might indicate on dislodged pollen clumps during microscopy workflow.

```{r scatterplot_pi_clsm_vs_callose_plug_index, echo=FALSE}
microscopy_df |>
        filter(!is.na(pi_clsm)) |>
        filter(!is.na(callose_plug_index)) |> 
        filter(pi_clsm > 0) |> # count() 41 flowers included in this subset
        ggplot(aes(x = pi_clsm, y = callose_plug_index)) +
        geom_point(size = 2, alpha = 0.6) +
        geom_smooth(
                method = "lm",
                se = FALSE,
                color = "black",
                linewidth = 0.8
                )


```

```{r distribution_of_callose_plug_index, echo=FALSE}
microscopy_df |>
        filter(!is.na(callose_plug_index)) |>
        filter(callose_plug_index > 0) |> # 44 flowers included in this subset
        ggplot(aes(x = callose_plug_index, fill = flower_treatment)) +
        geom_histogram(binwidth = 30, color = "black", alpha = 0.5, position = "stack") +
        # facet_wrap(~ management_type, ncol = 1) +
        labs(
                title = "Distribution of callose plug index",
                x = "Callose plug index,\n[number of blobs detected using Laplacian of Gaussian algorithm]",
                y = "Number of flowers"
        ) +
        microscopy_status_theme

```

### Developing tires for pollination and pollen performance phenotypes
```{r}
microscopy_df |>
  filter(
    !is.na(pi_clsm),
    !is.na(is_germination)
  ) |>
  mutate(
    ## ---- 1. Pollination delivery tiers ----
    pollination_tier = case_when(
      pi_clsm < 5                    ~ "None / noise (<5)",
      pi_clsm >= 5 & pi_clsm <= 110  ~ "Suboptimal (5вЂ“110)",
      pi_clsm > 110                  ~ "Saturating (>110)"
    ),

    ## ---- 2. Performance tiers ----
    performance_tier = case_when(
      is_germination == FALSE ~ "No germination",
      is_germination == TRUE & 
        callose_plug_index <= median(
          callose_plug_index[is_germination == TRUE],
          na.rm = TRUE
        ) ~ "Germination, low CPI",
      is_germination == TRUE & 
        callose_plug_index > median(
          callose_plug_index[is_germination == TRUE],
          na.rm = TRUE
        ) ~ "Germination, high CPI"
    )
  ) |>
  count(pollination_tier, performance_tier) |>
  group_by(pollination_tier) |>
  mutate(percent_within_tier = round(100 * n / sum(n), 1)) |>
  ungroup()
```

## Phenotyping stigma fungal colonization
To evaluate whether flower availability to visiting insects influences fungal colonization of cocoa stigmas, fungal hyphae presence was first summarized by flower treatment. Bagging represents a physical barrier limiting contact with insect vectors, and therefore provides a natural contrast for assessing environmentally mediated fungal exposure.

Fungal colonization was scored as a binary trait based on CLSM observations, indicating the presence or absence of hyphae on the stigma surface. Proportions were calculated within each `flower_treatment` to visualize relative frequencies of fungal colonization.

```{r fungal_colonization_by_flower_treatment, echo=FALSE, warning=FALSE, message=FALSE}
# ---- Fungal colonization by flower treatment (CLSM, dAB-only subset implicitly via NA filtering) ----
microscopy_df |>
        filter(!is.na(is_hyphae_present),
               !is.na(flower_treatment)) |>
        mutate(
                flower_treatment = factor(
                        flower_treatment,
                        levels = c("bagged_flower",
                                   "unbagged_flower")
                        ),
                hyphae_status = if_else(is_hyphae_present,
                                        "Hyphae present",
                                        "No hyphae"),
                hyphae_status = factor(hyphae_status,
                                       levels = c("Hyphae present",
                                                  "No hyphae")
                                       )
                ) |>
        count(flower_treatment, hyphae_status) |>
        group_by(flower_treatment) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(ggplot2::aes(x = flower_treatment, y = percent, fill = hyphae_status)) +
        geom_col(width = 0.7, color = "white", position = position_stack(reverse = TRUE)) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(labels = scales::percent_format(accuracy = 1), expand = c(0, 0)) +
        scale_fill_manual(name = "", values = pal_hyphae) +
        labs(
                title = "Microscopy detected fungal colonization of stigma by flower treatment",
                x = "Flower treatment",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme +
        theme(
                plot.title = element_text(hjust = 0.5),
                legend.position = "bottom"
                )

ggsave(
  filename = "results/fungal_stigma_colonization_by_flower_treatment.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)
```

To formally test whether flower treatment is associated with fungal colonization while accounting for non-independence among flowers sampled from the same tree, a generalized linear mixed-effects model (GLMM) with a binomial error distribution was fitted. Tree identity was included as a random intercept to control for repeated sampling within trees.


```{r glmm_hyphae_by_flower_treatment, echo=FALSE}
# ---- GLMM for fungal hyphae presence by flower treatment ----
# H0: The probability of detecting fungal hyphae is independent of flower treatment.
# H1: There is an association between flower treatment and fungal colonization on stigma.
glmm_hyphae_flower_treatment <- lme4::glmer(
        is_hyphae_present ~ flower_treatment + (1 | tree_id),
        data = microscopy_df |> dplyr::filter(!is.na(is_hyphae_present)),
        family = binomial
        )

summary(glmm_hyphae_flower_treatment)
```
To explore whether fungal hyphae presence exhibits additional structure across farms and management types, hyphae colonization occurrence was further summarized by `farm_id`. These visualizations are intended to provide descriptive insight into how fungal colonization varies across farms and management types.The limited sampling and limited number of farms per management category makes formal inference impossible.

NB! The sampling size for some farms (i.e. `ib`) is smaller compared to pollination analysis because only dAB stained samples were analysed.

```{r hyphae_presence_by_farm_both_flower_treatments, echo=FALSE, warning=FALSE, message=FALSE}
# ---- Fungal colonization by farm and management type (both bagged and unbagged flowers) ----

microscopy_df |>
        filter(!is.na(is_hyphae_present)) |>
        mutate(
                farm_id = factor(
                        farm_id,
                        levels = c("ib", "vr", "sa", "kk", "mt", "vi", "yb")
                        ),
                hyphae_status = if_else(
                        is_hyphae_present,
                        "Hyphae present",
                        "No hyphae"
                        ),
                hyphae_status = factor(
                        hyphae_status,
                        levels = c("Hyphae present", "No hyphae")
                        ),
                management_type = factor(
                        management_type,
                        levels = c(
                                "inside_forest",
                                "near_forest",
                                "agroforest",
                                "full_sun"
                                )
                        )
                ) |>
        count(management_type, farm_id, hyphae_status) |>
        group_by(farm_id) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = farm_id, y = percent, fill = hyphae_status)) +
        geom_col(
                width = 0.7,
                color = "white",
                position = position_stack(reverse = TRUE)
                ) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(
                labels = scales::percent_format(),
                expand = c(0, 0)
                ) +
        scale_fill_manual(name = "", values = pal_hyphae) +
        facet_grid(
                . ~ management_type,
                scales = "free_x",
                space  = "free_x",
                switch = "x"
                ) +
        labs(
                title = "Microscopy detected fungal colonization of stigma across farms",
                subtitle = "(both bagged and unbagged flowers)",
                x = "Farm ID and management type",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme +
        theme(
                strip.placement   = "outside",
                strip.background  = element_blank(),
                axis.text.x       = element_text(margin = margin(t = 5))
                )

ggsave(
  filename = "results/fungal_stigma_colonization_across_farm_all_treatments.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)
```
Because `bagged_flower` class has much lower fraction of flowers with fungal hyphae on stigma, patterns of fungal colonization were additionally visualized within `unbagged_flower` class only. This subset allows a more direct comparison among farms and management types under natural conditions, while avoiding satistical conclusions due to insufficient sampling effort.

```{r hyphae_presence_by_farm_only_open_flower_treatment, echo=FALSE, warning=FALSE, message=FALSE}
# ---- Fungal colonization by farm and management type (unbagged flowers only) ----
microscopy_df |>
        filter(!is.na(is_hyphae_present),
               flower_treatment == "unbagged_flower") |>
        mutate(
                farm_id = factor(
                        farm_id,
                        levels = c("ib", "vr", "sa", "kk", "mt", "vi", "yb")
                        ),
                hyphae_status = if_else(
                        is_hyphae_present,
                        "Hyphae present",
                        "No hyphae"
                        ),
                hyphae_status = factor(
                        hyphae_status,
                        levels = c("Hyphae present", "No hyphae")
                        ),
                management_type = factor(
                        management_type,
                        levels = c(
                                "inside_forest",
                                "near_forest",
                                "agroforest",
                                "full_sun"
                                )
                        )
                ) |>
        count(management_type, farm_id, hyphae_status) |>
        group_by(farm_id) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = farm_id, y = percent, fill = hyphae_status)) +
        geom_col(
                width = 0.7,
                color = "white",
                position = position_stack(reverse = TRUE)
                ) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(
                labels = scales::percent_format(),
                expand = c(0, 0)
                ) +
        scale_fill_manual(
                name = "",
                values = pal_hyphae
                ) +
        facet_grid(
                . ~ management_type,
                scales = "free_x",
                space  = "free_x",
                switch = "x"
                ) +
        labs(
                title = "Microscopy detected fungal colonization of stigma across farms",
                subtitle = "(unbagged flowers only)",
                x = "Farm ID and management type",
                y = "Percentage of flowers with hyphae assessment"
                ) +
        microscopy_status_theme +
        theme(
                strip.placement   = "outside",
                strip.background  = element_blank(),
                axis.text.x       = element_text(margin = margin(t = 5))
                )

ggsave(
  filename = "results/fungal_stigma_colonization_across_farms_unbagged_only.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)
```

Given that both pollen deposition and fungal colonization depend on exposure to the external environment and insect vectors, fungal presence was further examined in relation to pollination status to assess whether these two processes are directly associated or merely co-structured by `flower_treatment`.

Fungal hyphae presence was next summarized by pollination status as detected by CLSM. Fungal hyphae were slightly more frequently detected on pollinated stigmas. This visualization illustrates the apparent co-occurrence of pollen and fungal structures on stigmas, but does not account for experimental or hierarchical structure.

```{r hyphae_by_pollination_status_stacked, echo=FALSE, warning=FALSE, message=FALSE}

microscopy_df |>
        filter(
                !is.na(is_pollination_clsm),
                !is.na(is_hyphae_present)
                ) |>
        mutate(
                pollination_status = if_else(
                        is_pollination_clsm,
                        "Pollinated",
                        "Not pollinated"
                        ),
                pollination_status = factor(
                        pollination_status,
                        levels = c("Not pollinated", "Pollinated")
                        ),
                hyphae_status = if_else(
                        is_hyphae_present,
                        "Hyphae present",
                        "No hyphae"
                        ),
                hyphae_status = factor(
                        hyphae_status,
                        levels = c("Hyphae present", "No hyphae")
                        )
                ) |>
        count(pollination_status, hyphae_status) |>
        group_by(pollination_status) |>
        mutate(percent = n / sum(n)) |>
        ungroup() |>
        ggplot(aes(x = pollination_status, y = percent, fill = hyphae_status)) +
        geom_col(
                width = 0.7,
                color = "white",
                position = position_stack(reverse = TRUE)
                ) +
        geom_text(
                aes(label = n),
                position = position_stack(vjust = 0.5, reverse = TRUE),
                size = 3
                ) +
        scale_y_continuous(
                labels = scales::percent_format(accuracy = 1),
                expand = c(0, 0)
                ) +
        scale_fill_manual(
                values = pal_hyphae,
                name = ""
                ) +
        labs(
                title = "Microscopy detected fungal colonization of stigma by pollination status",
                x = "Microscopy verified pollination status",
                y = "Percentage of flowers"
                ) +
        microscopy_status_theme +
        theme(
                plot.title = element_text(hjust = 0.5),
                legend.position = "bottom"
                )

ggsave(
  filename = "results/fungal_stigma_colonization_by_pollination.png",
  plot = last_plot(),
  width = 7,
  height = 4,
  dpi = 300
)
```
To test whether fungal colonization is associated with pollination status while accounting for repeated sampling within trees, a GLMM was fitted with pollination status as the sole fixed effect and tree identity as a random intercept.

```{r}
# H0: Fungal hyphae presence is independent of pollination status.
# H1: The probability of detecting fungal hyphae differs between pollinated and non-pollinated stigmas.

glmm_hyphae_pollination <- glmer(
  is_hyphae_present ~ is_pollination_clsm + (1 | tree_id),
  data = microscopy_df |> 
    filter(
      !is.na(is_hyphae_present),
      !is.na(is_pollination_clsm)
    ),
  family = binomial
)

summary(glmm_hyphae_pollination)
```

Because pollination status and flower treatment are not independent unbagged flowers are more likely to receive pollenвЂ”a second model was fitted including both pollination status and flower treatment as fixed effects. This model allows disentangling the direct association between fungal colonization and pollen presence from their shared dependence on flower openness.

```{r}
glmm_hyphae_pollination_flower_treatment <- glmer(
        is_hyphae_present ~ is_pollination_clsm + flower_treatment + (1 | tree_id),
        data = microscopy_df |> 
                filter(
                        !is.na(is_hyphae_present),
                        !is.na(is_pollination_clsm),
                        !is.na(flower_treatment)
                        ),
        family = binomial
        )

summary(glmm_hyphae_pollination_flower_treatment)

```

Together, these models indicate that the fungal hyphae presence and pollination status is largely explained by flower treatment. Once `flower_treatment` is accounted for, pollination status itself does not independently predict fungal colonization. This suggests that both pollen deposition and fungal colonization are primarily structured by access to the external environment rather than being associated.

```{r}
glmm_hyphae_pollination_flower_treatment_open <- glmer(
        is_hyphae_present ~ is_pollination_clsm + (1 | tree_id),
        data = microscopy_df |> 
                filter(
                        !is.na(is_hyphae_present),
                        !is.na(is_pollination_clsm),
                        !is.na(flower_treatment),
                        flower_treatment == "unbagged_flower"
                        ),
        family = binomial
        )

summary(glmm_hyphae_pollination_flower_treatment_open)

```

